Distance Sampling in \proglang{R}

### 2015-11-28 paper.tex
ER finishes first run through of edits
some questions included for DLM included in italics in .rmd

\author{
David L Miller\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
}
\title{Distance Sampling in \proglang{R}}
\Keywords{distance sampling, abundance estimation, line transects, point transects, \proglang{R}}

@@

\Plainauthor{David L Miller, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R}
\Plainkeywords{distance sampling, abundance estimation, line transects, point transects, R}

@@

\Address{
    David L Miller\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
  URL: \url{http://converged.yt}\\~\\
      Jeffrey L Laake\\
  National Marine Mammal Laboratory\\
  Alaska Fisheries Science Center 7600 Sand Point Way N.E., Seattle, WA
  98115, USA\\
  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\

@@

Distance sampling (Buckland et al. 2001; Buckland et al. 2004)
encompases a suite of field methods and statistical models used to
estimate the abundance of biological populations. Distance sampling
field procedure can be thought of as an extension of plot sampling,
where we wish to take into account the decreasing probability of
detecting objects at increasing distance from the sampler. We do this by
building a model for detectability and use quantities calculated from
the detection function to adjust the observed counts to obtain an
estimate of abundance.

@@

For both points and lines, once the geometry of the sampler (see
``Detection functions'') has been taken into account, the histogram of
distances should show a decreasing number of observations with
increasing distance from the sampler. For line transects we expect
objects to be distributed unformly with respect to distance from the
line and what makes our histogram decrease is the detectability. For
point transects, we recognise that as distance from the point increases,
the area of the circle encompassed increases with distance squared;
hence the number of objects available to be detected is a linearly
increasing function of distance from the point.

Using this histogram we can crudely estimate the drop-off in
detectability by eye by tracing a line that approximates the tops of the
histogram bars -- this is the detection function. \pkg{Distance}
estimates the parameters for a fixed-form detection function using
maximum likelihood estimation. We address possible models in detail
below.

Figure \ref{fig:pointslines} shows examples of 500 individuals sampled
using line and point transects (left column) and their corresponding
histograms (right column).

@@

The line transect data is simulated data based on a survey of Antarctic
minke whales (\emph{Balaenoptera bonaerensis}). The data is simulated
from models fitted to data from the International Whaling Commission's
International Decade of Cetacean Research Southern Ocean Whale and
Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
surveys (Branch and Butterworth 2001). They consist of 99 observations
and include information on the effort expended and whether observations
were in one of two geographical strata (near or far from land).

@@

The point transect data set consists of 1485 observations of Amakihi
(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
between 1992 and 1995. The data include distances and three covariates
collected during the survey: observer ID (a three level factor), minutes
after sunrise (continuous) and hours after sunrise (a six level factor).
Data are analysed comprehensively in Marques et al. (2007).

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  In the simplest case, where one would simply like to estimate a
  detection function, all that is needed is a vector of distances.
\item
  To include additional covariates into the detection function (see
  ``Detection functions'') we need to use a \code{data.frame}. The
  data.frame contains a column called \code{distance} and additional
  named columns for covariates potentially useful in modelling
  detectability (for example \code{observer} or \code{seastate}). Some
  column names are reserved: \code{object} for an observation identifier
  (see ``Extensions''), \code{size} for group or cluster size (see
  ``Detection functions'' and ``Abundance and variance estimation''),
  \code{detected} for whether an observation was detected (see
  ``Extensions'') and the columns described in the next bullet.
\item
  If one would like to estimate abundance beyond area covered by
  expended survey effort, additional information is required. This
  consists of information about which transect and occasion the
  observation was made (the \code{Sample.Label}), a column named
  \code{Effort} which gives the effort associated with that sample (for
  lines their length and for points the number of times that point was
  visited), the stratum the sample was located in (this may have any
  name and may be from pre- or post-survey stratification, see
  ``Estimating abundance and variance'') and that stratum's area (which
  has the same name as the stratum column, appended with \code{.Area}).
  (We refer to this data format as ``flatfile'' where all information is
  contained in one table.)
\end{itemize}

As we will see in ``Extensions'', further information is also required
when we start using more complex models.

@@

As mentioned above, the detection function describes the relationship
between observed distances and probability of detection. The detection
function itself models the probability
\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
distance (either from a line or point) and \(\boldsymbol{\theta}\) is a
vector of parmeters to be estimated. Our goal is to estimate an
\emph{average probability of detection} (\(p\), average in the sense of
an average over the distances), so we must integrate out distance
(\(y\)) from the detection function: \[
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
\] where \(\pi(y)\) describes the distribution of objects with respect
to the sampler and \(\pi(x)=1/w\) for line transects and
\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
geometry of the sampler (Buckland et al. 2001, Chapter 3) (letting \(x\)
denote a perpendicular distance from a line and \(r\) denote radial
distance from a point).

It is important that the detection function accurately models
detectability near zero distance. We are less worried by its behaviour
further away from 0. To ensure that the model is not overly influence by
distances far from zero we truncate the distances beyond a given
distance \(w\) (known as the \emph{truncation distance}). Fitting to
distances at great distances from zero does not demonstrably improve the
precision of abundance estimates (Buckland et al. 2001, 103--8,
151--53).

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \emph{Shoulder}: we expect observers to be able to see objects near
  them, not just those directly infront of them. For this reason, we
  expect the detection function to be ``flat'' near zero distance.
\item
  \emph{Non-increasing}: we don't think that observers should be more
  likely to see things further away than those nearer to them. This
  usually indicates an issue with field procedure (that the distribution
  of objects with respect to the line, \(\pi(y)\) is not what we
  expect), so we do not want the detection function to model this.
\item
  \emph{Model robust}: models should be flexible enough to make many
  different shapes.
\item
  \emph{Pooling robust}: many factors can affect the probability of
  detection and it is not possible to measure all of these. We would
  like our models to be robust to us not including these factors.
\item
  \emph{Estimator efficiency}: we would like our models to have low
  variances, but only given they satisfy the other properties above
  (which, if they are satisfied, would give low bias).
\end{itemize}

Given these criteria, we can start to think about models for \(g\).

@@

There is a wide literature on possible formulations for the detection
function (Buckland 1992; Eidous 2005; Becker and Quang 2009; Giammarino
and Quatto 2014; Miller and Thomas 2015; Becker and Christ 2015).
\code{Distance} includes the most popular of these models. Here we'll
detail the most popular detection function approach: ``key function plus
adjustments'' (K+A).

@@

There are many factors that can affect the probability of detecting an
object. These include things like the observer, the vessel or platform
used, the sea state, weather conditions and time of day (to name but a
few). We assume that these variables affect detection only via the scale
of the detection function (and do not affect the shape).

Covariates can be included in this formulation by considering the scale
parameter from the half-normal or hazard-rate detection functions as a
linear model (on the exponential scale) of the (\(J\)) covariates
(\(\mathbf{z}\); a vector of length \(J\) for each observation): \[
\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
\] In the next section we'll discuss model selection.

Including covariates has an important implication for our calculation of
detectability. Since we don't know what the true distribution of the
covariates is we must calculate the probability of detection conditional
on the observed values of the covariates: \[
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
\] where \(\mathbf{z_i}\) is the vector of \(J\) covariates associated
with observation \(i\). So for covariate models, we are calculating a
value of ``average'' probability of detection (again the average is in
the sense of distance) per observation. There will be as many unique
values of \(p(\mathbf{z_i})\) as there are unique covariate combinations
in our data.

Another important consideration is that K+A models which include
covariates and one or more adjustments cannot be guaranteed to be
monotonic non-increasing for all covariate combinations, as we don't
have any model for the distribution of the covariates. For this reason,
we advise against using both adjustments and covariates in a detection
function (see Miller and Thomas 2015 for an example of when this can be
problematic).

@@

Note that \code{ds} will automatically select adjustment terms by AIC
and shows it's selection steps as it goes.

Figure \ref{fig:minkeamakihi} (top left plot) shows the result of
calling \code{plot} on the resulting model object. We can also call
\code{summary} on the model object to get summary information about the
fitted model (though we postpone this until the next section).

@@

Again, \code{ds} fits a model with cosine adjustments (the default) but
finds the AIC improvement to be insufficient to select the adjustment.

@@

By default \code{ds} assumes that the data given to it is line transect,
but we can switch to points using \code{transect="point"}. Including
covariates in the scale is via the \code{formula=~...} argument to
\code{ds}; a hazard-rate model for the amakihi that includes observer as
a covariate can be specified by (truncating at 82.5m, given in Marques
et al. 2007):

@@

As with the minke whale model, we can plot the resulting detection
function (Figure \ref{fig:minkeamakihi}, centre column). Since for the
amakihi we used covariates in the detection function, the plot will show
the detection function averaged over levels/values of the covariate.
Points on the plot indicate probability of detection for each
observation. For the \texttt{amakihi\_hr\_obs} model we see fairly clear
levels of the observer covariate in the points. Looking at the left
panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear
when adding minutes after sunrise as a covariate into the model:

@@

The bottom row of Figure \ref{fig:minkeamakihi} shows the probability
density functions for the amakihi models. These often are easier to
interpret than the detection functions, as for point data we must
rescale the histogram when plotting the detection function to take into
account the geometry of the point sampler.

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we
can use \code{summary} to give us some useful information about our
fitted model. For example for our hazard-rate model for the amakihi,
with observer as a covariate:

@@

This summary information includes various details of the data and model
specification, as well as the values of the coefficents (\(\beta_j\))
and their uncertainties, an ``average'' value for the detectability (see
``Estimating abundance and variance'' for details on how this is
calculated) and it's uncertainty. The final line gives an estimate of
abundance for the area covered by the survey (again, this will be
covered in detail in the next section).

@@

To judge goodness of fit for detection functions when exact distances
are used, we want to compare the cumulative distribution function (CDF)
and empirical distribution function (EDF) for the detection function via
a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the
probability of observing an object at a distance less than or equal to
some given value. The EDF gives the proportion of observations for which
the CDF is less than or equal to that of a given distance. In essence
we're asking ``is the number of observations up to a given distance in
line with what the model says they should be?'' (where `given values'
are the observed distances). As usual for Q-Q plots, ``good'' models
will have values close to the line \(y=x\), poor models will show more
deviations from that line.

We can inspect Q-Q plots visually, though for a large number of models
this can be tiresome and prone to subjective judgements. Instead we can
quantify the Q-Q plot's information using a Kolmogorov-Smirnov or
Cramer-von Mises test (Burnham et al. 2004). Both test to see if the
points from the EDF and CDF are from the same distribution. The
Kolmogorov-Smirnov uses the test statistic of the largest difference
between a point on the Q-Q plot and the line \(y=x\), whereas the
Cramer-von Mises test uses the sum of all the differences. As it takes
into account more information and is therefore more powerful, the
Cramer-von Mises is generally preferred. A significant result from
either tests indicates that the EDF and CDF do not come from the same
distribution (and therefore the model is does not fit the data well).

We can generate a Q-Q plot and test results using the \code{gof_ddf}
function. We can see the goodness of fit tests below for two models for
the amakihi data, first fitting a half-normal model without covariates
or adjustments (not setting \code{adjustment=NULL} will force \code{ds}
to fit a model with no adjustments), then calculating goodness of fit
for that model and our hazard-rate model with observer and minutes after
sunrise included:

@@

First, in order to estimate abundance in the covered area
(\(\hat{N}_\text{C}\)), we use the estimates of detection probability
(the \(\{p_i; i=1,\ldots,n\}\), above) in a Horvitz-Thompson like
estimator:

@@

where \(s_i\) are the sizes of the observed groups of objects, which is
equal to 1 if objects only occur singly (Borchers and Burnham 2004).
Thompson (2002) is the cannonical reference to this type of estimator;
intuitively, we can think of the estimates of detectability
(\(\hat{p}_i\)) as ``inflating'' the group sizes (\(s_i\)) to account
for incomplete detection -- we then sum to obtain the abundance
estimate. For models that do not include covariates, \(\hat{p}_i\) is
equal for all \(i\), so this is equivalent to summing the groups and
inflating that sum by the corresponding
\(\hat{p} (=\hat{p}_i \forall i)\).

@@

We can use the Horvitz-Thompson-like estimator to calculate the
``average'' detectability for models which include covariates. We can
consider what single detectability value would give the estimated
\(\hat{N}\) and therefore calculate: \[
\hat{P_a} = n/\hat{N}_\text{C}.
\] This can be a useful summary statistic and is included in the
\code{summary} output and the table produced by
\code{summarize_ds_models}.

@@

We may wish to calculate abundance estimates for some sub-regions of the
study region, we call these areas \emph{strata}. A stratum may be
defined by, for example, a different habitat type or by the sex of the
animal (or some combination) which may be of interest for biological or
management reasons. In order to calculate estimates for a given
stratification each observation must occur in a stratum which must be
labelled with a \code{Region.Label} and have a corresponding \code{Area}
(if we are using an animal characteristic like sex, we would have the
areas be the same but if we were using say forested vs.~wetland habitat
the areas of those strata would be different). Finally, we must also
know which stratum a given transect lies in.

As an example the minke whale data consists of two strata: \code{North}
and \code{South} relating to a stratum further away and nearer the
Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows
the two strata, along with observations and transect lines.

@@

Model uncertainty can be addressed using standard maximum likelihood
theory. We can invert the Hessian matrix of the likelihood to obtain a
variance-covariance matrix. We can then pre- and post-multiply this by
the derivatives of \(\hat{N}_\text{C}\) \[
\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
\] where the partial derivatives of \(\hat{N}_\text{C}\) are evaluated
at the MLE (\(\hat{\theta}\)) and \(\mathbf{H}\) is the first partial
Hessian (outer product of first derivatives of the log likelihood) for
numerical stability (Buckland et al. 2001, p 62). Note that although we
calculate uncertainty in \(\hat{N}_\text{C}\), we can trivially scale-up
to variance of \(\hat{N}\).

Sampling uncertainty can be characterised (rather than as simply \(n\))
by the \emph{encounter rate}: the number of objects per unit transect.
When covariates are not included in the detection function we can define
the encounter rate as \(n/L\) for line transects (where \(L\) is the
total line length) or \(n/T\) for point transects (where \(T\) is the
total number of visits to all points). When covariates are included in
the detection function, it is recommended that we substitute the \(n\)
in the previous expressions with the estimated abundance \(\hat{N}\) as
this will take into account the effects of the covariates.

For line transects, by default, \pkg{Distance} uses a variation of the
estimator ``R2'' from Fewster et al. (2009) replacing number of
observations per sample with the estimated abundance per sample (Innes
et al. 2002; F. F. C. Marques and Buckland 2003): \[
\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
\] where \(l_k\) are the lengths of the \(K\) transects (such that
\(L = \sum_{k=1}^K l_k\)). Whereas for points we use estimator ``P3''
from Fewster et al. (2009) but again replacing \(n\) by \(\hat{N}\) to
obtain the following estimator: \[
\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
\] where \(t_k\) is the number of visits to point \(k\) and
\(T = \sum_{k=1}^K t_k\).

@@

We combine these two sources of variance by noting that squared
coefficients of variation (approximately) add (Goodman 1960).

@@

Going back to the minke whale data, we have the information we require
to calculate \(A\) and \(a\) above, so we can estimate abundance and its
variance. When we supply data to \code{ds} in the ``flatfile'' format
given above, \code{ds} will automatically calculate abundance estimates
based on the survey setup in the data.

Having already fitted a model to the minke whale data, we can see the
results of the abundance estimation by simply looking at a model
summary:

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
David L Miller\\University of St Andrews \And Eric Rexstad\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
}
\title{Distance Sampling in \proglang{R}}
\Keywords{distance sampling, abundance estimation, line transects, point transects, \proglang{R}}

@@

\Plainauthor{David L Miller, Eric Rexstad, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R}
\Plainkeywords{distance sampling, abundance estimation, line transects, point transects, R}

@@

\Address{
    David L Miller\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
  URL: \url{http://converged.yt}\\~\\
      Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:Eric.Rexstad@st-andrews.ac.uk}{\nolinkurl{Eric.Rexstad@st-andrews.ac.uk}}\\

      Jeffrey L Laake\\
  National Marine Mammal Laboratory\\
  Alaska Fisheries Science Center 7600 Sand Point Way N.E., Seattle, WA
  98115, USA\\
  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\

@@

Distance sampling (Buckland et al. 2001; Buckland et al. 2004)
encompasses a suite of field methods and statistical models used to
estimate the abundance of biological populations. Distance sampling
field procedure can be thought of as an extension of plot sampling,
where we wish to take into account the decreasing probability of
detecting objects at increasing distance from the sampler. We do this by
building a model for detectability and use quantities calculated from
the detection function to adjust the observed counts to obtain an
estimate of abundance.

@@

For both points and lines, given the geometry of the sampler (see
``Detection functions''), the histogram of distances should show a
decreasing number of observations with increasing distance from the
sampler. For line transects we expect objects to be distributed
uniformly with respect to distance from the line with detectability
causing the number of detections to decrease. For point transects, we
recognise that as distance from the point increases, the area of the
circle encompassed increases with distance squared; hence the number of
objects available to be detected is a linearly increasing function of
distance from the point.

Using this histogram we can crudely estimate the drop-off in
detectability by eye by tracing a line that approximates the tops of the
histogram bars -- this is the detection function. \pkg{Distance}
estimates the parameters for a fixed-form detection function using
maximum likelihood estimation. We address possible models below.

Figure \ref{fig:pointslines} shows sampling of a population of 500
individuals using line and point transects (left column) and their
corresponding histograms (right column).

@@

The line transect data is based on a survey of Antarctic minke whales
(\emph{Balaenoptera bonaerensis}). The data are simulated from models
fitted to data from the International Whaling Commission's International
Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research
(IWC IDCR-SOWER) programme 1992-1993 austral summer surveys (Branch and
Butterworth 2001). They consist of 99 observations and include
geographical strata (near or distant from ice edge) and transect lengths
on the effort expended.

@@

The point transect data set consists of 1485 observations of Amakihi
(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
between 1992 and 1995. The data include distances and two covariates
collected during the survey: observer (a three level factor), time after
sunrise (transformed to minutes (continuous) or hours (factor)
covariates). Data are analysed comprehensively in Marques et al. (2007).

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  In the simplest case, where the objective is to estimate a detection
  function, all that is needed is a vector of distances.
\item
  To include additional covariates into the detection function (see
  ``Detection functions'') we use a \code{data.frame}. The data.frame
  contains a column called \code{distance} and additional named columns
  for covariates potentially useful in modelling detectability (for
  example \code{observer} or \code{seastate}). Some column names are
  reserved: \code{object} for an observation identifier (see
  ``Extensions''), \code{size} for group or cluster size (see
  ``Detection functions'' and ``Abundance and variance estimation''),
  \code{detected} for whether an observation was detected (see
  ``Extensions'') and the columns described in the next bullet.
\item
  To estimate abundance beyond area covered by expended survey effort,
  additional information is required. This consists of information about
  which transect and occasion the observation was made (the
  \code{Sample.Label}), a column named \code{Effort} which gives the
  effort associated with that sample (for lines their length and for
  points the number of times that point was visited), the stratum the
  sample was located in (this may have any name and may be from pre- or
  post-survey stratification, see ``Estimating abundance and variance'')
  and strata areas (which has the same name as the stratum column,
  appended with \code{.Area}). We refer to this data format as
  ``flatfile'' where all information is contained in one table.
\end{itemize}

As we will see in ``Extensions'', further information is also required
when using more complex models.

@@

The detection function describes the relationship between observed
distances and probability of detection. The detection function itself
models the probability
\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
distance (either from a line or point) and \(\boldsymbol{\theta}\) is a
vector of parameters to be estimated. Our goal is to estimate an
\emph{average probability of detection} (\(p\), average in the sense of
an average over distances), so we must integrate out distance (\(y\))
from the detection function: \[
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
\] where \(\pi(y)\) describes the distribution of objects with respect
to the sampler and \(\pi(x)=1/w\) for line transects and
\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
geometry of the sampler (Buckland et al. 2001, Chapter 3) (letting \(x\)
denote a perpendicular distance from a line and \(r\) denote radial
distance from a point).

It is crucial that the detection function accurately models
detectability at small distances. We are less worried by its behaviour
further away from 0. To ensure that the model is not overly influence by
distances far from zero, we truncate the distances beyond a given
distance \(w\) (known as the \emph{truncation distance}). Fitting to
distances at great distances from zero does not demonstrably improve the
precision of abundance estimates (Buckland et al. 2001, 103--8,
151--53).

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \emph{Shoulder}: we expect observers to be able to see objects near
  them, not just those directly adjacent to them. For this reason, we
  expect the detection function to be horizontal near zero distance.
\item
  \emph{Non-increasing}: we do not think that observers should be more
  likely to see things distant from the transect than those near the
  transect. If this occurs, it usually indicates an issue with field
  procedure (that the distribution of objects with respect to the line,
  \(\pi(y)\) is not what we expect), so we do not want the detection
  function to model this.
\item
  \emph{Model robust}: models should be flexible enough to make many
  different shapes.
\item
  \emph{Pooling robust}: many factors can affect the probability of
  detection and it is not possible to measure all of these. We would
  like models to produce unbiased results without inclusion of these
  factors.
\item
  \emph{Estimator efficiency}: we would like models to have low
  variances, but only given they satisfy the other properties above
  (which, if they are satisfied, would give low bias).
\end{itemize}

Given these criteria, we can formulate models for \(g\).

@@

There is a wide literature on possible formulations for the detection
function (Buckland 1992; Eidous 2005; Becker and Quang 2009; Giammarino
and Quatto 2014; Miller and Thomas 2015; Becker and Christ 2015).
\code{Distance} includes the most popular of these models. Here we will
detail the most popular detection function approach: ``key function plus
adjustments'' (K+A).

@@

There are many factors that can affect the probability of detecting an
object. These include things like the observer, the vessel or platform
used, the sea state, weather conditions and time of day. We assume that
these variables affect detection only via the scale of the detection
function (and do not affect the shape).

Covariates can be included in this formulation by considering the scale
parameter from the half-normal or hazard-rate detection functions as a
linear model (on the exponential scale) of the (\(J\)) covariates
(\(\mathbf{z}\); a vector of length \(J\) for each observation): \[
\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
\] In the next section we will discuss model selection.

Including covariates has an important implication for our calculation of
detectability. Because we do not know what the true distribution of the
covariates is we must calculate the probability of detection conditional
on the observed values of the covariates: \[
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
\] where \(\mathbf{z_i}\) is the vector of \(J\) covariates associated
with observation \(i\). For covariate models, we are calculating a value
of ``average'' probability of detection (average in the sense of
distance) per observation. There will be as many unique values of
\(p(\mathbf{z_i})\) as there are unique covariate combinations in our
data.

K+A models that include covariates and one or more adjustments cannot be
guaranteed to be monotonic non-increasing for all covariate
combinations, as we do not have any model for the distribution of the
covariates. For this reason, we advise against using both adjustments
and covariates in a detection function (see Miller and Thomas 2015 for
an example of when this can be problematic).

@@

Note that \code{ds} will automatically select adjustment terms by AIC
and shows the selection steps.

Figure \ref{fig:minkeamakihi} (top left plot) shows the result of
calling \code{plot} on the resulting model object. We can also call
\code{summary} on the model object to get summary information about the
fitted model (we postpone this to the next section).

@@

\code{ds} fits a model with cosine adjustments (the default) but finds
the AIC improvement to be insufficient to select the adjustment.

@@

By default \code{ds} assumes the data given to it is line transect, but
we can switch to points using \code{transect="point"}. Including
covariates in the scale is via the \code{formula=~...} argument to
\code{ds}; a hazard-rate model for the amakihi that includes observer as
a covariate can be specified by (truncating sightings at 82.5m, given in
Marques et al. 2007):

@@

As with the minke whale model, we can plot the resulting detection
function (Figure \ref{fig:minkeamakihi}, centre column). Because the
amakihi model included covariates, the plot shows the detection function
averaged over levels/values of the covariate. Points on the plot
indicate probability of detection for each observation. For the
\texttt{amakihi\_hr\_obs} model we see fairly clear levels of the
observer covariate in the points. Looking at the left panel of Figure
\ref{fig:minkeamakihi}, we can see this is less clear when adding
minutes after sunrise as a covariate into the model:

@@

The bottom row of Figure \ref{fig:minkeamakihi} shows the probability
density functions for the amakihi models. Probability density function
plots give a better sense of fit of the model than the detection
functions, as for point data we must rescale the histogram when plotting
the detection function to take into account the geometry of the point
sampler.

** ER suggests the g(x) plots for the amakihi be deleted **

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we
can use \code{summary} to give useful information about our fitted
model. For example for our hazard-rate model for the amakihi, with
observer as a covariate:

@@

This summary information includes details of the data and model
specification, as well as the values of the coefficients (\(\beta_j\))
and their uncertainties, an ``average'' value for the detectability (see
``Estimating abundance and variance'' for details on how this is
calculated) and its uncertainty. The final line gives an estimate of
abundance for the area covered by the survey (covered in the next
section).

@@

To judge goodness of fit for detection functions when exact distances
are used, we want to compare the cumulative distribution function (CDF)
and empirical distribution function (EDF) for the detection function via
a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the
probability of observing an object at a distance less than or equal to
some given value. The EDF gives the proportion of observations for which
the CDF is less than or equal to that of a given distance. This can be
interpreted as assessing whether the number of observations up to a
given distance in line with what the model says they should be (where
`given values' are the observed distances). As usual for Q-Q plots,
``good'' models will have values close to the line \(y=x\), poor models
will show more deviations from that line.

We can inspect Q-Q plots visually, though this is prone to subjective
judgments. Instead we can quantify the Q-Q plot's information using a
Kolmogorov-Smirnov or Cramer-von Mises test (Burnham et al. 2004). Both
test whether points from the EDF and CDF are from the same distribution.
The Kolmogorov-Smirnov uses the test statistic of the largest difference
between a point on the Q-Q plot and the line \(y=x\), whereas the
Cramer-von Mises test uses the sum of all the differences. As it takes
into account more information and is therefore more powerful, the
Cramer-von Mises is generally preferred. A significant result from
either tests indicates that the EDF and CDF do not come from the same
distribution (and therefore the model is does not fit the data well).

We can generate a Q-Q plot and test results using the \code{gof_ddf}
function. We can see the goodness of fit tests below for two models for
the amakihi data (Figure \ref{amakihi-qq}), first fitting a half-normal
model without covariates or adjustments (not setting
\code{adjustment=NULL} will force \code{ds} to fit a model with no
adjustments), then calculating goodness of fit for that model and our
hazard-rate model with observer and minutes after sunrise included:

@@

First, to estimate abundance in the covered area (\(\hat{N}_\text{C}\)),
we use the estimates of detection probability (the
\(\{p_i; i=1,\ldots,n\}\), above) in a Horvitz-Thompson like estimator:

@@

where \(s_i\) are the sizes of the observed groups of objects, which is
equal to 1 if objects only occur singly (Borchers and Burnham 2004).
Thompson (2002) is the canonical reference to this type of estimator;
intuitively, we can think of the estimates of detectability
(\(\hat{p}_i\)) as ``inflating'' the group sizes (\(s_i\)) to account
for incomplete detection -- we then sum over all \(n\) detections to
obtain the abundance estimate. For models that do not include
covariates, \(\hat{p}_i\) is equal for all \(i\), so this is equivalent
to summing the groups and inflating that sum by the corresponding
\(\hat{p} (=\hat{p}_i \forall i)\).

@@

\emph{have L and K been previously defined?}

We can use the Horvitz-Thompson-like estimator to calculate the
``average'' detectability for models which include covariates. We can
consider what single detectability value would give the estimated
\(\hat{N}\) and therefore calculate: \[
\hat{P_a} = n/\hat{N}_\text{C}.
\] This can be a useful summary statistic and is included in the
\code{summary} output and the table produced by
\code{summarize_ds_models}. \emph{anything else to say about this, seems
to mess up folks, including me}

@@

We may wish to calculate abundance estimates for some sub-regions of the
study region, we call these areas \emph{strata}. For example, strata may
be defined by habitat types or animal gender (or some combination) which
may be of interest for biological or management reasons. To calculate
estimates for a given stratification each observation must occur in a
stratum which must be labelled with a \code{Region.Label} and have a
corresponding \code{Area} (if we are using an animal characteristic like
gender, we would have the areas be the same but if we were using say
forested vs.~wetland habitat the areas of those strata would be
different). Finally, we must also know the stratum in which transects
lie.

As an example the minke whale data consists of two strata: \code{North}
and \code{South} relating to a stratum further away and nearer the
Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows
the two strata, along with observations and transect lines. \emph{figure
number messed up}

@@

Model uncertainty can be addressed using standard maximum likelihood
theory. We can invert the Hessian matrix of the likelihood to obtain a
variance-covariance matrix. We can then pre- and post-multiply this by
the derivatives of \(\hat{N}_\text{C}\) with respect to the parameters
of the detection function \[
\widehat{\text{Var}}_\text{model}\left( \hat{N}_\text{C}\right) = \left(\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}\right)^\text{T} \left(\hat{\mathbf{H}}(\hat{\boldsymbol{\theta}})^{-1} \right)\frac{\partial \hat{N}_\text{C}}{\partial\hat{\boldsymbol{\theta}}}
\] where the partial derivatives of \(\hat{N}_\text{C}\) are evaluated
at the MLE (\(\hat{\theta}\)) and \(\mathbf{H}\) is the first partial
Hessian (outer product of first derivatives of the log likelihood) for
numerical stability (Buckland et al. 2001, p 62). Note that although we
calculate uncertainty in \(\hat{N}_\text{C}\), we can trivially scale-up
to variance of \(\hat{N}\).

Sampling uncertainty can be characterised (rather than as simply \(n\))
by the \emph{encounter rate}: the number of objects per unit transect.
When covariates are not included in the detection function we can define
the encounter rate as \(n/L\) for line transects (where \(L\) is the
total line length) or \(n/T\) for point transects (where \(T\) is the
total number of visits to all points) \emph{earlier eqn uses notation
\(K\) for number of points}. When covariates are included in the
detection function, it is recommended that we substitute the \(n\) in
the previous expressions with the estimated abundance \(\hat{N}\) as
this will take into account the effects of the covariates.

For line transects, by default, \pkg{Distance} uses a variation of the
estimator ``R2'' from Fewster et al. (2009) replacing number of
observations per sample with the estimated abundance per sample (Innes
et al. 2002; F. F. C. Marques and Buckland 2003): \[
\widehat{\text{Var}}_{\text{encounter},R2} = \frac{K}{L^2(K-1)} \sum_{k=1}^{K} l_k^2 \left( \frac{N_k}{l_k} - \frac{N}{L}\right)^2,
\] where \(l_k\) are the lengths of the \(K\) transects (such that
\(L = \sum_{k=1}^K l_k\)). Whereas for points we use estimator ``P3''
from Fewster et al. (2009) but again replacing \(n\) by \(\hat{N}\) to
obtain the following estimator: \[
\widehat{\text{Var}}_{\text{encounter},P3} = \frac{1}{T(K-1)} \sum_{k=1}^{K} t_k \left( \frac{N_k}{t_k} - \frac{N}{T}\right)^2,
\] where \(t_k\) is the number of visits to point \(k\) and
\(T = \sum_{k=1}^K t_k\). \emph{this messes with the \(K\) notation even
further}

@@

We combine detection function variance with encounter rate variance by
use of the \(\delta\)-method approximation (Goodman 1960).

@@

Returning to the minke whale data, we have the necessary information to
calculate \(A\) and \(a\) above, so we can estimate abundance and its
variance. When we supply data to \code{ds} in the ``flatfile'' format
given above, \code{ds} will automatically calculate abundance estimates
based on the survey information in the data.

Having already fitted a model to the minke whale data, we can see the
results of the abundance estimation by viewing the model summary:
